Modal type theory based on the intuitionistic epistemic logic

### 2018-01-18 article.tex
erased rubbish

  Modal intutionistic epistemic logic IEL was proposed by S. Artemov and T. Proropopescu \cite{Artemov}. IEL
  provides the epistimology and the theory of knowledge as based on BHK-semantics of intuitionistic logic.
  $IEL^{-}$  is a variant of IEL, that corresponds to intuitionistic belief.
  Informally, $\textbf{K}A$ denotes that $A$ is verified intuitionistically.

@@

    2) $\textbf{K}(A \to B) \to (\textbf{K}A \to \textbf{K}B)$ (normality);

    3) $A \to \textbf{K}A$ (co-reflection);

@@

  $\lambda K$ consists of the rules for simply typed lambda-calculus and special typing rules for
  lifting types into the applicative functor $\textbf{K}$. We assume that our type system will
  axiomatize the simplest case of computation with effects with one container. We provide proof-theoretical
  view on
  this kind of computations in functional programming and prove strong normalization and confluence.

@@

  Let $\mathbb{V}$ be the set of variables. The set $\Lambda_{\textbf{K}}$ of terms is defined by the grammar:

$\begin{array}{lll}
& \Lambda_{\textbf{K}} ::= \mathbb{V} \: | \:  (\lambda \Lambda.\Lambda_{\textbf{K}}) \: | \: (\Lambda_{\textbf{K}}\Lambda_{\textbf{K}}) \: | \: (\Lambda_{\textbf{K}} , \Lambda_{\textbf{K}}) \: | \: (\pi_1 \Lambda_{\textbf{K}}) \: | \: (\pi_2 \Lambda_{\textbf{K}}) \: | & \\
& \quad\quad\quad\quad\quad\quad\quad\quad\quad\quad\quad\quad\quad\quad\quad\quad ({\bf pure \: } \: \Lambda_{\textbf{K}}) \: | \: ({\bf let \:\: pure} \:\: \Lambda_{\textbf{K}} = \Lambda_{\textbf{K}} \:\: {\bf in} \:\: \Lambda_{\textbf{K}})
\end{array}$

@@

  Let $\mathbb{T}$ be the set of atomic types. The set $\mathbb{T}_{\textbf{K}}$ of types with applicative
  functor
  $\textbf{K}$ is generated by the grammar:
  \begin{equation}
    \mathbb{T}_{\textbf{K}} ::= \mathbb{T} \: | \: (\mathbb{T}_{\textbf{K}} \to \mathbb{T}_{\textbf{K}}) \: |
    \:
    (\mathbb{T}_{\textbf{K}} \times \mathbb{T}_{\textbf{K}}) \: | \: (\textbf{K}\mathbb{T}_{\textbf{K}})
  \end{equation}
  \end{defin}

@@

  \begin{defin} Modal typed lambda calculus $\lambda \textbf{K}$ based on NIEL$^{-}_{\land, \to}$:

@@

    \begin{prooftree}
      \AxiomC{$\Gamma \vdash \vec{M} : {\bf K} \vec{A}$}
      \AxiomC{$\vec{x} : \vec{A} \vdash M : B$}
      \RightLabel{$\text{let}_{\bf K}$}
      \BinaryInfC{$\Gamma \vdash {\bf let \: pure \:} \vec{x} = \vec{M} {\: \bf in \: } M : {\bf K} B$}
    \end{prooftree}
  \end{tabular}
  \end{minipage}%

@@

  $\textbf{K}_I$-typing rule is the same as $\bigcirc$-introduction in lax logic (also known as monadic
  metalanguage \cite{Lax}) and in typed lambda-calculus which is derived by proof-assignment for lax-logic
  proofs.
  $\textbf{K}_I$ allows to inject an object of type $\alpha$ into the functor. $\textbf{K}_I$  reflects the
  Haskell method {\bf pure} for Applicative class. It plays the same role as the {\bf return} method in Monad class.

  $\text{let}_{\bf K}$ is similar to $\Box_I$-rule in typed lambda calculus for intuitionistic normal modal logic {\bf IK}, which is described in \cite{ModalK}.

@@

  5) ${\bf pure \:} ((\lambda x. M) N) \rightarrow_{\beta} {\bf pure \: } (M [x := N])$;

  6) ${\bf pure \:} (\pi_i \langle M_1, M_2 \rangle) \rightarrow_{\beta} {\bf pure \:} M_i$, where $i \in \{ 1, 2 \}$.

  7) $\begin{array}{llll}
  & {\bf pure \:} ({\bf let \: pure \:} \langle \vec{x}, y, \vec{z} \rangle = \langle \vec{M}, {\bf let \: pure \: } \vec{w} = \vec{N} {\bf \: in \: } Q, \vec{P} \rangle {\: in \:} R) \rightarrow_{\beta} & \\
  & {\bf pure \:} ({\bf let \: pure \:} \langle \vec{x}, \vec{w}, \vec{z} \rangle = \langle \vec{M}, \vec{N}, \vec{P} \rangle \: { \bf in } \: R [y := Q])
  \end{array}$

  8) $\lambda x. f x \rightarrow_{\eta} f$;

  9) $\langle \pi_1 P, \pi_2 P \rangle \rightarrow_{\eta} P$;

@@

  12) ${\bf pure \:} (\lambda x. f x) \rightarrow_{\eta} {\bf pure \:} f$;

  13) ${\bf pure \:} (\langle \pi_1 P, \pi_2 P \rangle) \rightarrow_{\eta} {\bf pure \:} P$;

  14) ${\bf pure \:} ({\bf let \: pure \:} x = M {\: \bf in \: } x) \rightarrow_{\eta} {\bf pure \:} M$;

  15) ${\bf pure \:} ({\bf let \: pure \:} \underline{\quad} = \underline{\quad} {\: \bf in \:} N) \rightarrow_{\eta} { \bf pure \:} ({\bf pure \:} N)$.
  \end{defin}

@@

  \begin{defin} The domain of a context $\Gamma$:

  Let $\Gamma = \{ x_1 : A_1, ..., x_n : A_n \}$. Then the domain of $\Gamma$, or $dom(\Gamma)$, is a set $\{  x_1, ..., x_n\}$.

  \begin{lemma}

  If $\Gamma \vdash M : A$, then $FV(M) \subseteq dom(\Gamma)$

  Induction on the derivation of $\Gamma \vdash M : A$.

  \begin{lemma} Generation for $\lambda \textbf{K}$.

  1) $\Gamma \vdash {\bf pure \: } M : \textbf{K}\alpha$ implies that $\Gamma \vdash M : \alpha$;

  2) $\Gamma \vdash {\bf let \: pure} \: \vec{N} = \vec{M} \: {\bf in} \: M : {\bf K} B$ implies that $\Gamma \vdash \vec{M} : {\bf K} \vec{A}$ and
$\vec{N} : \vec{A} \vdash M : B$.
  \end{lemma}

Induction on the derivation of $\Gamma \vdash {\bf pure \: } M : \textbf{K}\alpha$ and $\Gamma \vdash {\bf let \: pure} \: \vec{N} = \vec{M} \: {\bf in} \: M : {\bf K} B$ respectively.
  \end{proof}

  The next one lemma allows that weakening structural rule is admissable.

  \begin{lemma} Weakening for $\lambda \textbf{K}$.

  Let $\Gamma \vdash M : A$ and $\Gamma \subseteq \Delta$, then $\Delta \vdash M : A$.

  Induction on derivation of $\Gamma \vdash M : A$. Let us assume $\Gamma \subseteq \Delta$.

  1) Let $\Gamma \vdash x : A$, such that $\Gamma = \Delta, x : A$ and $\Theta \subseteq \Gamma$.
  Let $\Sigma = \Theta \setminus \Gamma$, or, which is the same, $\Sigma = \Theta \setminus \Delta, x :
  A$, then $\Sigma, \Delta, x : A \vdash x : A$, or, $\Theta \vdash x : A$.

  2) Let $\Gamma \vdash {\bf pure \: } M : \textbf{K}A$ and $\Gamma \subseteq \Theta$.

  By generation $\Gamma \vdash M : A$

  By hypothesis, $\Theta \vdash M : A$, so $\Theta \vdash {\bf pure \: } M : \textbf{K}A$ by applying
  $\textbf{K}_I$-rule.

  3) Let $\Gamma \vdash {\bf let \: pure \:} \vec{x} = \vec{M} {\: \bf in \:} N : {\bf K} B$ and $\Gamma \subseteq \Theta$.

  By generation $\Gamma \vdash \vec{M} : {\bf K} \vec{A}$ and $\vec{x} : \vec{A} \vdash N : B$.

  By assumption $\Theta \vdash \vec{M} : {\bf K} \vec{A}$.

  Hence $\Theta \vdash {\bf let \: pure \:} \vec{x} = \vec{M} {\: \bf in \:} N : {\bf K} B$.

  \begin{lemma} Considering for $\lambda \textbf{K}$.

  If $\Gamma \vdash M : \alpha$, then $\Gamma \uparrow FV(M) \vdash M : \alpha$, where $\Gamma \uparrow
  FV(M)$ is a subcontext of $\Gamma$, such that $dom(\Gamma \uparrow FV(M)) = dom(\Gamma) \cap FV(M)$.

  1) Let $\Gamma \vdash x : A$, where $\Gamma = \Delta, x : A$, $x \in \mathbb{V}$.

  $FV(x) = \{ x\}$, then $dom(\Gamma) \cap \{ x\} = \{ x\}. $ So $(\Delta, x : A) \uparrow FV(x) = \{ x
  : A\}$, then $x : A \vdash x : A$ by axiom.

  2) Let $\Gamma \vdash {\bf pure \: } M : {\bf K} A$.

  By generation, $\Gamma \vdash M : A$ and $\Gamma \uparrow FV(M) \vdash M : A$ by hypothesis.

  So $\Gamma \uparrow FV(M) \vdash {\bf pure \: } M : {\bf K} A$ by ${\bf K}_I$.

  3) Let $\Gamma \vdash {\bf let \: pure \:} \vec{x} = \vec{M} {\: \bf in \:} N : {\bf K} B$.

By generation, $\Gamma \vdash \vec{M} : {\bf K} \vec{A}$ and $\vec{x} : \vec{A} \vdash N : B$.

By assumption, $\Gamma \uparrow FV(\vec{M}) \vdash \vec{M} : {\bf } \vec{A}$.

By $\text{let}_{\bf K}$, $\Gamma \uparrow FV(\vec{M}) \vdash {\bf let \: pure \:} \vec{x} = \vec{M} {\: \bf in \:} N : {\bf K} B$

  \begin{lemma}

  If $\Gamma, x : A \vdash M : B$  and $\Gamma \vdash N : A$, then $\Gamma \vdash (M [x := N]) : B$

  1) Let $\Gamma, x : A \vdash {\bf pure \: } M : \textbf{K}B$ and $\Gamma \vdash N : A$.

  By generation, $\Gamma, x : A \vdash M : B$.

  By assumption, $\Gamma \vdash (M [x := N]) : B$

  Then $\Gamma \vdash {\bf pure \: } (M [x := N]) : \textbf{K}B$ by $\textbf{K}_I$.

  But ${\bf pure \: } (M [x := N]) = ({\bf pure \: } M [x := N]) $ by substitution definition, so
  $\Gamma \vdash ({\bf pure \: } M [x := N]) : \textbf{K}B$

  2) Let $\Gamma, y : A \vdash {\bf let \: pure \:} \vec{x} = \vec{M} {\: \bf in \:} N : {\bf K} B$ and $\Gamma \vdash N : A$.

  By generation, $\Gamma, y : A \vdash \vec{M} : {\bf K} \vec{A}$ and $\vec{x} : \vec{A} \vdash N : B$.

  By hypothesis, $\Gamma \vdash \vec{M}[x := N] : {\bf K} \vec{A}$.

  Hence $\Gamma \vdash {\bf let \: pure \:} \vec{x} = \vec{M}[x := N] {\: \bf in \:} N : {\bf K} B$.

  \begin{theorem} Subject reduction

  i) Let $\Gamma \vdash M : A$ and $M \twoheadrightarrow_{\beta} N$, then $\Gamma \vdash N : A$

  ii) Let $\Gamma \vdash M : A$ and $M \twoheadrightarrow_{\eta} N$, then $\Gamma \vdash N : A$
  \end{theorem}

We consider only modal $\beta$-reduction rules. The general statement for $\twoheadrightarrow_{\beta}$ follows from transitivity of multi-step $\beta$-reduction.

  i) For multistep $\beta$-reduction:

1) Let $\Gamma \vdash {\bf let \: pure \:} \langle \vec{x}, y, \vec{z} \rangle = \langle \vec{M}, {\bf let \: pure \: } \vec{w} = \vec{N} { \: \bf in \: } Q, \vec{P} \rangle { \: \bf in \: } R : {\bf K }B$

By generation we have $\Gamma \vdash \vec{M} : {\bf K}\vec{A_1}$, $\Gamma \vdash {\bf let \: pure \: } \vec{w} = \vec{N} {\: \bf in \:} Q : {\bf K}\vec{A_2}$, $\Gamma \vdash \vec{P} : {\bf K}\vec{A_3}$
and $\vec{x} : \vec{A_1}, y : A_2, \vec{z} : \vec{A_3} \vdash R : B$.

If $\Gamma \vdash {\bf let \: pure \: } \vec{w} = \vec{N} {\bf \: in \: } Q : {\bf K}\vec{A_2}$, then

$\Gamma \vdash \vec{N} : {\bf K}\vec{A_4}$ and $\vec{w} : \vec{A_4} \vdash Q : A_2$.

Then $\vec{x} : \vec{A_1}, \vec{w} : \vec{A_4}, \vec{z} : \vec{A_3} \vdash R [y := Q] : B$ by substitution lemma and weakening.

Hence $\Gamma \vdash {\bf let \: pure \:} \langle \vec{x}, \vec{w}, \vec{z} \rangle = \langle \vec{M}, \vec{N}, \vec{P} \rangle \: { \bf in } \: R [y := Q] : {\bf K} B$
by $let_{{\bf K}}$.

2) Let $\Gamma \vdash {\bf pure } ((\lambda x. M) N) : {\bf K} B$.

By generation $\Gamma \vdash (\lambda x. M) N : B$, but $\Gamma \vdash M [x := N] : B$, then, by ${\bf K}_I$, $\Gamma \vdash {\bf pure \:} (M [x := N]) : {\bf K} B$.

3) Let $\Gamma \vdash {\bf pure } (\pi_i \langle M_1, M_2 \rangle) : {\bf K} A_i$, where $i \in \{ 1,2 \}$.

By generation $\Gamma \vdash \pi_i \langle M_1, M_2 \rangle : A_i$ and $\Gamma \vdash M_i : A_i$.

Hence $\Gamma \vdash {\bf pure \: } M_i : { \bf K } A_i$ by ${\bf K}_I$.

4) Let $\Gamma \vdash {\bf let \: pure \:} (\langle \vec{x}, y, \vec{z} \rangle = \langle \vec{M}, {\bf let \: pure \: } \vec{w} = \vec{N} {\bf \: in \: } Q, \vec{P} \rangle { \: \bf in \: } R) : {\bf K }^2 B$.

By generation $\Gamma \vdash {\bf let \: pure \:} \langle \vec{x}, y, \vec{z} \rangle = \langle \vec{M}, {\bf let \: pure \: } \vec{w} = \vec{N} {\bf \: in \: } Q, \vec{P} \rangle { \: \bf in \: } R : {\bf K }B$,

hence $\Gamma \vdash {\bf let \: pure \:} \langle \vec{x}, \vec{w}, \vec{z} \rangle = \langle \vec{M}, \vec{N}, \vec{P} \rangle \: { \bf in } \: R [y := Q] : {\bf K} B$ by the first case.

So $\Gamma \vdash {\bf pure \:} ({\bf let \: pure \:} \langle \vec{x}, \vec{w}, \vec{z} \rangle = \langle \vec{M}, \vec{N}, \vec{P} \rangle \: { \bf in } \: R [y := Q]) : {\bf K}^2 B$
by ${\bf K}_I$.

ii) For multistep $\eta$-reduction:

  1) Let $\vdash {\bf let \: pure \:} \underline{\quad} = \underline{\quad} {\: \bf in \:} N : {\bf K} A$.

  Then by generation $\vdash N : A$, so $\vdash {\bf pure \:} N : {\bf K} A$ by ${\bf K}_I$.

  2) Let $\Gamma \vdash {\bf let \: pure \:} x = M {\: \bf in \: } x : {\bf K} A$.

  By generation $\Gamma \vdash M : {\bf K} A$ and $x : A \vdash x : A$, hence $\Gamma \vdash M : {\bf K} A$.

  3) Let $\Gamma \vdash {\bf pure \:} (\lambda x. f x) : {\bf K}(A \to B)$.

  By generation $\Gamma \vdash \lambda x. f x : A \to B$, so $\Gamma \vdash f : A \to B$, then $\Gamma \vdash {\bf pure \:} f : {\bf K}(A \to B)$ by ${\bf K}_I$.

  4) Let $\Gamma \vdash {\bf pure \:} (\langle \pi_1 P, \pi_2 P \rangle) : {\bf K}(A \times B)$.

  By generation $\Gamma \vdash \langle \pi_1 P, \pi_2 P \rangle : A \times B$, then $\Gamma \vdash P : A \times B$.

  By ${\bf K}_I$, $\Gamma \vdash {\bf pure \:} P : {\bf K}(A \times B)$.

  5) $\Gamma \vdash {\bf pure \:} ({\bf let \: pure \:} x = M {\: \bf in \: } x) : {\bf K}^2 A$.

  Then $\Gamma \vdash M : {\bf K}^2 A$ and $x : {\bf K} A \vdash x : {\bf K} A$, so $\Gamma \vdash M : {\bf K}^2 A$.

  6) Let $\vdash {\bf pure \:} ({\bf let \: pure \:} \underline{\quad} = \underline{\quad} {\: \bf in \:} N) : {\bf K}^2 A$.

  By generation ${\bf let \: pure \:} \underline{\quad} = \underline{\quad} {\: \bf in \:} N : {\bf K}A$, so $\vdash N : A$, then $\vdash {\bf pure \:} N : {\bf K}$.

@@

  2) $\ast_{A, B} \mathcal{F}A \otimes_2 \mathcal{F}B \to \mathcal{F}(A \otimes_1 B)$

>>>>

  Modal intutionistic epistemic logic IEL was proposed by S. Artemov and T. Proropopescu \cite{Artemov}. IEL
  provides the epistimology and the theory of knowledge as based on BHK-semantics of intuitionistic logic.
  IEL$^{-}$  is a variant of IEL, that corresponds to intuitionistic belief.
  Informally, ${\bf K}A$ denotes that $A$ is verified intuitionistically.

@@

    2) ${\bf K}(A \to B) \to ({\bf K}A \to {\bf K}B)$ (normality);

    3) $A \to {\bf K}A$ (co-reflection);

@@

  $\lambda_{{\bf K}}$ consists of the rules for simply typed lambda-calculus and special typing rules for
  lifting types into the applicative functor $\textbf{K}$. We assume that our type system will
  axiomatize the simplest case of computation with effects with one container. We provide proof-theoretical
  view on
  this kind of computations in functional programming and prove strong normalization and confluence.

@@

  Let $\mathbb{V}$ be the set of variables. The set $\Lambda_{{\bf K}}$ of terms is defined by the grammar:

$\begin{array}{lll}
& \Lambda_{{\bf K}} ::= \mathbb{V} \: | \:  (\lambda \Lambda.\Lambda_{{\bf K}}) \: | \: (\Lambda_{{\bf K}}\Lambda_{{\bf K}}) \: | \: (\Lambda_{{\bf K}} , \Lambda_{{\bf K}}) \: | \: (\pi_1 \Lambda_{{\bf K}}) \: | \: (\pi_2 \Lambda_{{\bf K}}) \: | & \\
& \quad\quad\quad\quad\quad\quad\quad\quad\quad\quad\quad\quad\quad\quad\quad\quad ({\bf pure \: } \: \Lambda_{{\bf K}}) \: | \: ({\bf let \:\: pure} \:\: \Lambda_{{\bf K}} = \Lambda_{{\bf K}} \:\: {\bf in} \:\: \Lambda_{{\bf K}})
\end{array}$

@@

  Let $\mathbb{T}$ be the set of atomic types. The set $\mathbb{T}_{{\bf K}}$ of types with applicative
  functor ${\bf K}$ is generated by the grammar:
  \begin{equation}
    \mathbb{T}_{{\bf K}} ::= \mathbb{T} \: | \: (\mathbb{T}_{{\bf K}} \to \mathbb{T}_{{\bf K}}) \: |
    \: (\mathbb{T}_{{\bf K}} \times \mathbb{T}_{{\bf K}}) \: | \: ({\bf K}\mathbb{T}_{{\bf K}})
  \end{equation}
  \end{defin}

@@

  \begin{defin} Modal typed lambda calculus $\lambda_{{\bf K}}$ based on NIEL$^{-}_{\land, \to}$:

@@

    \begin{prooftree}
      \AxiomC{$\Gamma \vdash \vec{M} : {\bf K} \vec{A}$}
      \AxiomC{$\vec{x} : \vec{A} \vdash M : B$}
      \RightLabel{$\text{let}_{{\bf K}}$}
      \BinaryInfC{$\Gamma \vdash {\bf let \: pure \:} \vec{x} = \vec{M} {\: \bf in \: } M : {\bf K} B$}
    \end{prooftree}
  \end{tabular}
  \end{minipage}%

@@

  ${\bf K}_I$-typing rule is the same as $\bigcirc$-introduction in lax logic (also known as monadic
  metalanguage \cite{Lax}) and in typed lambda-calculus which is derived by proof-assignment for lax-logic
  proofs.
  ${\bf K}_I$ allows to inject an object of type $\alpha$ into the functor. ${\bf K}_I$  reflects the
  Haskell method {\bf pure} for Applicative class. It plays the same role as the {\bf return} method in Monad class.

  $\text{let}_{{\bf K}}$ is the same as the $\Box$-rule in typed lambda calculus for intuitionistic normal modal logic {\bf IK}, which is described in \cite{ModalK}.

  In fact, our calculus is the extention of typed lambda calculus for {\bf IK} with typing rule appropriate to co-reflection.

@@

  5) $M \rightarrow_{\beta} N \Rightarrow {\bf pure \: M} \rightarrow_{\beta} {\bf pure \: N}$

  6) $\lambda x. f x \rightarrow_{\eta} f$;

  7) $\langle \pi_1 P, \pi_2 P \rangle \rightarrow_{\eta} P$;

@@

  12) $M \rightarrow_{\beta} N \Rightarrow {\bf pure \: M} \rightarrow_{\eta} {\bf pure \: N}$
  \end{defin}

@@

@@

  2) $\ast_{A, B} : \mathcal{F}A \otimes_2 \mathcal{F}B \to \mathcal{F}(A \otimes_1 B)$
